---
title: 'US Natality Data: Causal Inference Case Study'
subtitle: 'Analysis'
output:
  html_document:
    df_print: paged
---

```{r echo=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE, comment=NA)

library(ggplot2)   # plot
library(dplyr)     # tables
library(readr)     # reading
library(stringr)   # strings
library(purrr)     # fp
library(tidyr)     # remove_na
library(forcats)   # factor recode
library(glmnet)    # lasso
library(grf)       # random forests
library(janitor)   # remove_constant columns
library(corrplot)  # plot correlation matrix
library(sandwich)  # robust covariance matrix
#library(sufrep)   # sufficient representations
source("make_encoder.R") # modified sufficient representations

rm(list = ls())    # reset environment
source("utils.R")  # import helpers

# The seed is set before every randomized operation throughout so results
# are consistent when rerunning cells.
```

This notebook implements a case study of some causal inference methods on an
observational dataset, the
[Vital Statistics Natality Birth Data](https://www.nber.org/research/data/vital-statistics-natality-birth-data).
Specifically, we attempt to estimate the causal effect of a mother being a heavy
smoker (vs. non-smoker) on infant birth weight, using data from 100,000 hospital
births.

```{r}
year <- 2002
treatment <- "smoking_heavy"
```

Load data and models.

```{r}
# load data with original factors
natl_fname <- paste0("models/natl/natl", year, ".", treatment, ".rds")
natl <- read_rds(natl_fname)

# load data with factors encoded
natl_fname <- paste0("models/natl_rep/natl", year, ".", treatment, ".rds")
natl_rep <- read_rds(natl_fname)

# load model
train_fname <- paste0("models/causal_forests/natl", year, treatment, ".rds")
tau.forest <- read_rds(train_fname)
```

# Averate Treatment Effect

A first-pass naive attempt at making a causal estimate could simply look at a
difference in means and ignore possible confounding. This would only be
justified if the assignment of smoking / not smoking status was independent of
the potential outcomes, i.e. the birth weight in either case. This assumption is
almost certainly false, as numerous medical and socio-economic confounders could
effect both smoking and health outcomes such as birth weight.

```{r}
difference_in_means(natl_rep, Yvar="dbirwt", Wvar="smoking_heavy")
```

Instead, we use a powerful non-parametric estimator: the causal forest (TODO
CITE). When we control for our selected covariates using this model, a smaller
magnitude average effect is estimated. This suggests that confounding is at play
and that it likely contributes to the difference-in-means overestimating the
causal effect.

```{r}
average_treatment_effect(tau.forest)
```

Plotting the estimated smoking probabilities against mean outcomes for non-smoking
mothers, we see that among them, those who are more likely to smoke (in the
counterfactual sense) have lower birth weights. This suggests that confounding
factors that make mothers more likely to smoke during pregnancy also cause lower
birth weights. This lines up with the fact that the difference-in-means
estimator overestimated the magnitude of the effect when compared with the
causal forest.

```{r}
tibble(prop.hat = tau.forest$W.hat,
       mean.hat = tau.forest$Y.hat,
       W = tau.forest$W.orig) %>%
  # filter to untreated (non-smokers)
  filter(W == 0) %>%
  rename(mean0.hat = mean.hat) %>%
  # plot
  ggplot(aes(x = prop.hat, y = mean0.hat)) +
  geom_point(alpha = 1/20) +
  geom_quantile(formula = y ~ x, 
                quantiles = c(0.5), 
                show.legend=TRUE) +
  coord_cartesian(xlim = c(0, 0.4), ylim = c(3000, 3800))
```

And the same plot with estimated treatment propensities scaled by their empirical 
cumulative distribution.

```{r}
tibble(prop.hat.quantile = ecdf(tau.forest$W.hat)(tau.forest$W.hat),
       mean.hat = tau.forest$Y.hat,
       W = tau.forest$W.orig) %>%
  # filter to untreated (non-smokers)
  filter(W == 0) %>% 
  rename(mean0.hat = mean.hat) %>%
  # plot
  ggplot(aes(x = prop.hat.quantile, y = mean0.hat)) +
  geom_point(alpha = 1/20) +
  geom_quantile(formula = y ~ x, 
                quantiles = c(0.5), 
                show.legend=TRUE) +
  coord_cartesian(xlim = c(0, 1), ylim = c(3000, 4000))
```

# Averate Treatment Effect on the Treated

Since estimated treatment probabilities go very low (<1%) due to the fact that
smoking is quite uncommon during pregnancy, treatment effects for some controls
may not be well identified. For this reason, it can be helpful to look at the
ATT (average treatment effect on the treated). Note that the point estimate is
almost identical but the standard error is smaller (i.e. the effect is better
identified.) TODO check / change

```{r}
average_treatment_effect(tau.forest, target.sample = "treated")
```

# Treatment Effect Heterogeneity 

The calibration test indicates we are picking up meaningful heterogeneity.

```{r}
test_calibration(tau.forest)
```

The distribution of predicted causal effects is solidly negative which is what
would be expected.

```{r}
hist(tau.forest$predictions)
```

Computing the best linear projections of the 

```{r}
set.seed(157)
tau.blp.coefs <- as.list(colnames(tau.forest$X.orig)) %>%
  # calculate best linear projections onto single variables
  map(~ best_linear_projection(tau.forest, 
                               A = tau.forest$X.orig[,.x]))
names(tau.blp.coefs) <- colnames(tau.forest$X.orig)

tau.blp <- tau.blp.coefs %>%
  # filter constant vars w no coefficient
  keep(~ dim(.x)[1] == 2) %>%
  # extract into tibble rows
  imap_dfr(~ tibble(var = .y,
                    est = .x["A1", "Estimate"],
                    std = .x["A1", "Std. Error"],
                    tst = .x["A1", "t value"],
                    prb = .x["A1", "Pr(>|t|)"]))
tau.blp %>%
  arrange(prb)
```

## Live Birth Order and Mother's Age

The three variables with the most significant associations with heterogeneity in
the treatment effect are `dlivord` (what number live birth), `nlbnl` (number of
previous live births now living), and `dmage` (mother's age). The first two are
extremely correlated since `dlivord = nlbnl + nlbnd + 1` where most previous
live births are still alive.

```{r}
natl %>%
  sample_n(10000) %>%
  ggplot(aes(x = nlbnl, y = dlivord)) +
  geom_jitter(alpha = 1/10)
```

`dlivord` and `dmage` also correlated, although less highly.

```{r}
natl %>%
  sample_n(10000) %>%
  ggplot(aes(x = dmage, y = dlivord)) +
  geom_jitter(alpha = 1/10, width = 0.5, heigh = 0.5)
```

Since these variables have such a high correlation with heterogeneity in the 
treatment effect, we can see that the estimated average effect changes
significantly (in the practical sense) over their ranges.

First `dlivord`.

```{r}
dlivord.values <- 1:5
dlivord.table <- natl_rep$dlivord %>% 
  table() %>% 
  as_tibble() %>%
  rename(value = ".") %>%
  mutate(value = as.numeric(value))

as.list(dlivord.values) %>%
  map_dfr(~ average_treatment_effect(tau.forest,
                                     subset = natl$dlivord == .x,
                                     target.sample = "treated")) %>%
  mutate(value = dlivord.values, 
         .before = estimate)  %>%
  inner_join(dlivord.table, by = "value") %>%
  select(value, n, estimate, std.err)
```

Then `dmage`.

```{r}
quantiles <- 5
probs <- (0:quantiles) / quantiles

dmage.quantiles <- natl_rep$dmage %>% 
  cut(quantile(., probs = probs, na.rm = TRUE), 
      include.lowest = TRUE)

dmage.table <- dmage.quantiles %>% 
  table() %>% 
  as_tibble() %>% 
  rename(value = ".")

as.list(levels(dmage.quantiles)) %>%
  map_dfr(~ average_treatment_effect(tau.forest, 
                                     subset = dmage.quantiles == .x)) %>%
  mutate(value = levels(dmage.quantiles), 
         .before = estimate)  %>%
  inner_join(dmage.table, by = "value") %>%
  select(value, n, estimate, std.err)
```

The ATE estimates do not change monotonically with either `dlivord` or `dmage`.
This could be due to our model producing poor estimates (the standard errors
are quite large compared to the differences in value), or there could be real
nonlinearities in their relationships.

#### Population Size of City of Residence

Since this variable is a 5 level factor, we split by level.

```{r}
# citrspop <- Xmod[,"citrspop"] %>% 
#   table() %>%
#   as_tibble() %>%
#   rename(code = ".") %>%
#   mutate(code = as.integer(code)) %>%
#   mutate(value = case_when(
#     code == 0 ~ "> 1M",
#     code == 1 ~ "[500K, 1M]",
#     code == 2 ~ "[250K, 500K]",
#     code == 3 ~ "[100K, 250K]",
#     code == 9 ~ "< 100K"
#   ), .after = code)
```

```{r}
# as.list(citrspop$code) %>%
#   map_dfr(~ average_treatment_effect(tau_bin.forest, 
#                                  target.sample = "treated",
#                                  subset = Xmod[,"citrspop"] == .x)) %>%
#   mutate(value = citrspop$value, 
#          .before = estimate) %>%
#   inner_join(citrspop, by = "value") %>%
#   select(code, value, n, estimate, std.err)
```
